---
title: "Data Wrangling Example"
output:
  html_document:
    highlight: tango
    theme: cerulean
    toc: yes
    toc_float: yes
    toc_depth: 3
    keep_md: true
---

```{r setup, include=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr) # for text string stuff
library(forcats) # for factor stuff
library(knitr) # for markdown options
library(kableExtra) # makes tables nicer
```


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")

show_df <- function(df){
  modify_if(df, .p = is_character, ~ replace(.x, is.na(.x), "")) %>%
  kable(format = "markdown") %>% 
  print()
}
```
Importing and tidy some ficitious data

NB: These results are very much made up!


## Import data

From sheet number 1 of the spreadsheet

(Can use the sheet "name" instead of sheet number)

```{r}
DATA_FILE_NAME <- "ih-trial_results_20171020.xlsx"

dat <- readxl::read_excel(DATA_FILE_NAME, sheet = 1)
```
  

```{r, results='asis', echo=FALSE}
dat %>% show_df()
```


## Fill down the 'gappy' columns

```{r}
dat <- dat %>% 
  fill(sex, age, treatment)
```
```{r, results='asis', echo=FALSE}
dat %>% show_df()
```



## Make the subject ID nice

```{r}
dat <- dat %>% 
  mutate(subject = paste0("A", str_pad(subject, 3, "left", "0")))
```

```{r, results='asis', echo=FALSE}
dat %>% show_df()
```


## Clean the sex/status column

### Tally sex spellings

This doesn't change the dataframe - just prints a sumary table

```{r,fig.width=4}
dat %>%
  group_by(sex) %>%
  tally() %>% 
  kable(format = "html", caption = "Recorded sexes") %>% 
  kable_styling(full_width = FALSE) # looks better
```

### Fix sex spellings

```{r}
dat <- dat %>%
  mutate(sex = case_when(sex == "female nneutered" ~ "fn",
                         sex == "male entire" ~ "me",
                         sex == "MN" ~ "mn",
                         TRUE ~ sex))
```

```{r, results='asis', echo=FALSE}
dat %>% show_df()
```


## Separate sex/status into 2 columns

```{r}
dat <- dat %>% 
  tidyr::separate(sex, c("sex", "neuter_status"), 1)
```

```{r, results='asis', echo=FALSE}
dat %>% show_df()
```


## Make ages in months into years

```{r, size=0.1}
dat <- dat %>%
  mutate(age = case_when(
                    str_detect(age, "month") ~ parse_number(age) / 12,
                    TRUE ~ parse_number(age)))
```

```{r, results='asis', echo=FALSE}
dat %>% show_df()
```


## Gather results and add week column

```{r}
dat <- dat %>%
  tidyr::gather("week", "glucose", `week 1`:`week 4`) %>% 
  mutate(week = parse_number(week))
```

```{r, results='asis', echo=FALSE}
dat %>% show_df()
```


## Check glucose for obvious typos

```{r, fig.height=3}
ggplot(dat, aes(x = glucose)) +
  geom_histogram(binwidth = 5) +
  coord_cartesian(ylim = c(0, 10)) + 
  labs(title = "Quick histogram of all glucoses",
       subtitle = "NB: y axis truncated")
```

### Make a table of unusual results

This doesn't change the dataframe - just makes a table

```{r}
dat %>% 
  dplyr::filter(glucose > 50) %>% 
  dplyr::select(subject, week, rep, glucose) %>% 
  kable(format = "html", caption = "Unusually high glucose results") %>% 
  kable_styling(full_width = FALSE) # looks better
```

### Now fix unusually high results

(Lets assume we went back to the paper records)

```{r}

dat <- dat %>%
  mutate(glucose = case_when(subject == "A006" &
                             week == 1 &
                             rep == 2 ~         6.04,
                           
                           subject == "A012" &
                             week == 2 &
                             rep == 2 ~         7.76,
                           
                           subject == "A012" &
                             week == 3 &
                             rep == 1 ~         7.23,
                           
                           subject == "A003" &
                             week == 4 &
                             rep == 2 ~         4.08,
                           
                           subject == "A004" &
                             week == 4 &
                             rep == 1 ~         7.37,
                           
                           TRUE ~               glucose))
  
```


```{r, results='asis', echo=FALSE}
dat %>% show_df()
```

## Create a new column for when glucose > 5

```{r}
dat <- dat %>%
  mutate(glucose_category = case_when(glucose > 5.0 ~ "high",
                                  glucose <= 5.0 ~ "normal",
                                  NA ~ NA_character_,
                                  TRUE ~ "other"))
```

```{r, results='asis', echo=FALSE}
dat %>% show_df()
```

## Summary proportions

### Group the data by subject (ID) into a new dataframe

Then count high glucose measurements (and total number of measurements)

```{r}
dat_summary <- dat %>%  
  group_by(subject) %>% 
  summarise(n_high = sum(glucose_category == "high"),
            n = sum(!is.na(glucose_category))) %>% 
  ungroup()
```

```{r, results='asis', echo=FALSE}
dat_summary %>% show_df()
```

### And calculate proportion of readings that are high

```{r}
dat_summary <- dat_summary %>% 
  mutate(proportion_high = n_high / n)
```

```{r, results='asis', echo=FALSE}
dat_summary %>% show_df()
```

### Add a column with propotion as a round %

```{r}
dat_summary <- dat_summary %>% 
  mutate(percentage_high = round(proportion_high * 100, digits = 0))
```

```{r, results='asis', echo=FALSE}
dat_summary %>% show_df()
```

